---
title: "Midterm2024"
author: "Ruiming Song"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
library(styler)  # styler can automatically style code for you. Use it via the Addins menu at the top, right above all the tabs.

# Path to the helper.R file, relative to where your R markdown file is. (The '..' means 'go up a directory')
source("../helper.R")   
```





## Short Awnswer 

### Question 1 (6 pts)
Recall your Bayesian data analysis mantra. 

Part A)
Write it out in word format:

Bayesian data analysis refers to the posterior probability equals to the Prior probability times likelihood and divide by the sum of likelihood.


Part B)
Write it out in probability statement format (i.e. Pr(B|A) etc):

$$Pr(A|B)=\frac{Pr(B|A)Pr(A)}{Pr(B)}
$$




### Question 2 (6 pts)
What are the 3 sections that we can use to mathematically describe a Bayesian model?  

Prior distribution, likelihood function and post distribution



### Question 3 (6 pts)
Your friend shows you a dataset from their experiment in astral zenobiology. They have 20 independent observations of the mental life-force of virtual lifeforms, and they insist that due to their knowledge of the system, the appropriate distribution for the observations is the Sakanaesse distribution, defined as $S(y | \mu,\theta)$ (here we are defining the probability (or likelihood) of seeing a value $y$ given two parameters $\mu$ and $\theta$). 

The data you are given are in a table where column $y$ is the observed value, so $y_i$ is the value for individual observation $i$. 

Part A) Write out the likelihood formula for these data.

The likelihood is
$$ y_i \sim \text{Sakanaesse}(\mu, \theta) $$



Part B) Write out the likelihood formula on the log scale,
$$log( \text{Sakanaesse}(\mu, \sigma) )$$




## Analysis Challenge (82 points)

For this midterm exercise we will use data from this paper:  https://onlinelibrary.wiley.com/doi/10.1111/oik.07674 . We will use only a portion of their data, but our analysis will involve similar models to the ones discussed in the paper. You are free to read the paper or look to it for modeling inspiration, but you can complete this entire exercise without looking at the paper. 


In this study, clown fish were observed in breeding groups associated with sea anemones. In the dataset, each row is an observation.

The dataframe has 236 observations of 61 unique fish groups (labeled by Anemone_ID). Anemone size is taken to represent the food-richness of the area the fish live in, so anemone size might influence fish health and therefore fish reproductive output. 

The dataset includes a treatment, which is that some of the groups of fish were fed additional food. The column "FedIndex" is 1 if the fish were not fed, and 2 if they were fed. The treatment was performed in the middle of the season, so that some clutches of eggs were produced before the treatment, and others were produced after the treatment. The column "PostTreatment" is 1 if the clutch was laid before the treatment and 2 if it was after. Note that fish in the FedIndex=1 category were never fed additional food, even if PostTreatment=2. 

First clear your working environment and load the data.
```{r}
rm(list=ls())

load("ClownFishData.RData")
```

You now have an object called "data" in your environment. Take a few minutes to inspect it.




### (1) Standardize and plot
We will be using anemone_area as a predictor. Since it is a continuous variable, it's a good idea to standardize it. Name the standardized version of this column "AA".


Make a figure showing the relationship between anemone area and number of eggs laid.
```{r}
data$AA <- standardize(data$anemone_area)

ggplot( data , aes(x=AA ,y=Eggs_Laid )) +geom_point()
```

### (2) Plotting a prior
You will construct a linear regression model for the number of eggs laid with the anemone area (standardized) as the predictor. The model is

$$
\mathrm{Eggs\_Laid} \sim \mathrm{Normal(\mu,\sigma)}\\
\mu = a + b * \mathrm{AA} \\
a \sim \mathrm{Normal(400,150)} \\
b \sim \mathrm{Normal(0,200)} \\
\sigma \sim \mathrm{Exponential(0.01)} \\
$$
Plot the prior with the data. Explain what makes this a reasonable prior.
```{r}

m <- quap(
  alist(
    Eggs_Laid ~ dnorm(mu, sigma),
    mu <- a + b * AA,
    a ~ dnorm(400, 150),
    b ~ dnorm(0, 200),
    sigma ~ dexp(0.01)
  ),
  data = data,
  start = list(a = 250, b = 65, sigma = 140)
)

prior <- extract.prior(m)
seq_AA <- seq(from = -2, to = 2, length.out = 30)
samples_prior_m <- link_df(m, data = list(AA = seq_AA), post = prior)

ggplot(data = samples_prior_m, aes(x = AA, y = mu)) +
  geom_point(alpha = .05)

```
The high intercept around 400 gives a wide range.The exponential prior is relatively low. The slope b is also wide. Thus, it is a reasonable prior.
### (3) Linear regression model
Construct a *quap* model for the data and use `precis` to summarize the output. 

The *quap* is a bit sensitive with these data, so it helps to specify the initial conditions. Use `start= list(a=250,b=65,sigma=140)` as an option for *quap*.
```{r}
m <- quap(
  alist(
    Eggs_Laid ~ dnorm(mu, sigma),
    mu <- a + b * AA,
    a ~ dnorm(400, 150),
    b ~ dnorm(0, 200),
    sigma ~ dexp(0.01)
  ),
  data = data,
  start = list(a = 250, b = 65, sigma = 140)
)

# Summarize the model
precis(m)
```
### (4) Explain the precis output
The a is the intercept from the posterior distribution with a mean 267.47 and standard deviation of the posterior distribution of the intercept 8.98. 89% confident that the true value of the intercept lies within the range from 253.12 to 281.82. The b is the average value of the slope for anemone size with a mean 68.15 and standard deviation of the posterior distribution of the intercept 9.00. 89% confident that the true value of the anemone size within the range from 53.76 to 82.54. The sigma is the mean of the standard deviation of the residuals with a value of 138.15 and a standard deviation of sigma's posterior distribution of 6.33. 89% confident that the sigma lies within the range from 128.03 to 148.26.These result indicates that anemone size is positively associated with the number of eggs laid with a relatively low uncertainty based on the narraw credible interval. Howevever, the large standard deviation indicates the high variability in the number of eggs laid. Such variability may be caused by other factors.


### (5) Plot the linear regression lines
Use `link_df` to create samples of the *quap* fit. To do this, create a dataframe with evenly spaced out values of AA. Plot the mu values on a graph with the data. 
```{r}
AA_seq <- seq(from = min(data$AA), to = max(data$AA), length.out = 100)
AA_df <- data.frame(AA = AA_seq)
# Assuming 'm' is your model fit from quap
# Assuming 'm' is your model fit from quap
posterior_predictions <- link_df(m, data = AA_df)

ggplot(data, aes(x = AA, y = Eggs_Laid)) +
  geom_point() +  # This will plot the actual data points
  geom_line(data = posterior_predictions, aes(x = AA, y = mu), color = "blue", alpha = 0.4)  # This will plot the fitted regression lines

```

### (6) List one "big world" explanations for why the data show more variability than the model fit does.
The clownfish population is highly complex and variable, which means the the number of eggs laid may not only vary with the anemone area. Other factors will also affect the number of eggs laid such as female size, temperature changes, etc. Thus,  the data show more variability than the model fit does.

### (7) Including the treatment effect 
Split the data into two datasets, one for fish who received the treatment, and the other for fish who did not receive the treatment. 

You will analyze each of these datasets with a multivariate model that builds on your prior model. In addition to the effect of anemone size, include an effect based on whether or not the clutch of eggs was laid before or after the feeding occurred (remember PostTreatment=1 before feeding, and 2 after feeding). 

For each dataset, perform the quap fit and use `precis` to summarize the results.
```{r}
# Assuming `FedIndex` is the column indicating whether fish were fed or not
treated <- subset(data, FedIndex == 2)
untreated <- subset(data, FedIndex == 1)
m_treated <- quap(
   alist(
    Eggs_Laid ~ dnorm(mu, sigma),
    mu <- a + b_AA * AA + b_PT * PostTreatment,
    a ~ dnorm(0, 10),
    b_AA ~ dnorm(0, 10),
    b_PT ~ dnorm(0, 10),
    sigma ~ dexp(1)
  ), 
  data = treated
)
precis(m_treated)
```
```{r}
m_untreated <- quap(
   alist(
    Eggs_Laid ~ dnorm(mu, sigma),
    mu <- a + b_AA * AA + b_PT * PostTreatment,
    a ~ dnorm(0, 10),
    b_AA ~ dnorm(0, 10),
    b_PT ~ dnorm(0, 10),
    sigma ~ dexp(1)
  ), 
  data = untreated
)

# Summarize the models

precis(m_untreated)

```
### (8) Interpret the quap fits. What can you say about how the two datasets differ from each other in terms of their response to anemone size and to pre/post treatment?
For  the anemone size, both treated and untreated groups show a significant positive effect of anemone size on the number of eggs laid, but the untreated group is slightly stronger. As for the PT post treatment, the treated groups show a higher effect on more egg laid.Thus, the treatment and anemone size  both afect the eggs laid data.
